The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2018-12-01 Paper.Rmd
made at start at revising the introduction

---
title: "The importance of a complete statistical description of dependence between variables for ecology and related fields"
author: "Shyamolina Ghosh, Lawrence W Sheppard, Mark T. Holder, Terrance E. Loecke, Daniel C. Reuman"
date: "February 18, 2018"
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

@@

We used datasets that pertain to multiple fields and levels of organization within or related to ecology, including environmental, species-trait, phenological, population, community, and ecosystem functioning data. Copula structure describing the relationship between atmospheric weather variables such as X and Y has been examined previously in the meteorological literature (e.g., REF). We therefore examined environmental variables from the soil instead, using the Rapid Carbon Assessment database (RaCA; @Wills2014).
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil nitrogen (Mg C or N per hectare of soil surface) 
<!--***DAN: Shya, this used to say Mg C, but what is Mg? Do you mean mg for miligram? I have written it out, please check and correct if necessary. Also, I added per m^2 of soil surface, please check.
***Shya: Mg means mega gram, So an "Mg C" is a ton of Carbon - the standard SI unit when it comes to talking about soil Carbon measurement.-->
<!--***DAN: Can you check this please? I do not think it can be right. The values on Fig. 3a range from close to 0 to over 3000. 3000 would be 3000 Mg C. I do not think a cubic meter of soil even weighs that much, including all the carbon and non-carbon components.
***Shya: I checked : it should be megagram per hectare, Wills2014 reference is for stock of Carbon, not for nitrogen, Prof. loecke told me nitrogen data is not published yet. He screened locations which had both soil C and N stocks after omiting NAs.-->
from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. X for sampling locations).
<!--***DAN: Shya, please add a sup mat figure showing the sampling locations and a link to it here in place of the X
***Shya: once Prof. Loecke give me the lattitude and longitudes, I will do that.-->
Measurments were from the central pedon of a soil core, to 1 m depth.
<!--***DAN: Terry to check the wording and accuracy here-->
Species-trait data were average species basal metabolic rate (KJ per hour) and body mass (gram) data for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 35 years from 1976 to 2010. These time series were computed from the the Rothamsted Insect Survey suction-trap dataset [@Macaulay1988, @Harrington2014]. The first of our two population-level datasets was also derived from the RIS suction-trap data, and comprised total counts of aphids trapped for the same locations, species, and years; these are indicators of flying aphid abundance. Data pre-processing for the aphid data was the same as that of @Sheppard2016,
but we exlcuded one of the 11 sites they used so as to include only sites with at least 30 years data.  Our second population dataset comprised average annual abundance estimates for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 plankton taxa (Table \ref{SM-tab_plankton_info}) for the 56 years from 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. These data were obtained from the Continuous Plankton Recorder (CPR) dataset maintained by the Sir Alistair Hardy Foundation for Ocean Science (SAHFOS), now part of the Marine Biological Association of the UK. Data pre-processing was similar but not identical to that of @Sheppard2017; see Appendix \ref{SM-Data_details} for details.
Community-level data were plant aboveground biomass and Shannon's diversity index for X plots, each 9m by 9m, maintained at Cedar Creek Ecosystem Science Reserve and sampled in the year 2000 [@Tilman2001; @Tilman2006]. 
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or water surface and the troposphere, measured at 13 locations at daily to weekly intervals from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, Ohio [@Holland1999; @Jarecke2016]. Each included location was measured on at least 50 dates. Sampling locations spanned a hydrologic gradient from upland to seasonal wetland.
See Appendix \ref{SM-Data_details} for additional data details.

@@

# Concepts for \hyperref[Intro]{\textbf{(Q1)}}: Statistical tools for detecting copula structure \label{Methods}

Question \hyperref[Intro]{\textbf{(Q1)}} from the \nameref{Intro} was answered via statistical analysis of the
data described above, using both model selection and nonparametric approaches.

>>>>

---
author: "Shyamolina Ghosh, Lawrence W Sheppard, Mark T. Holder, Terrance E. Loecke, Daniel C. Reuman"
date: "February 18, 2018"
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

All branches of ecology and related areas study relationships among environmental and biological 
variables. However, commonly used correlation- and regression-based approaches to studying 
such relationships are limited - it is well understood that common statistical
approaches may provide only a very partial description of a relationship. 
For instance, datasets showing markedly different 
relationships may nevertheless have the same correlation coefficient value:  
Fig. \ref{fig_cop_pedag1}A-B show two such datasets. The panel A 
(respectively, B) variables of Fig. \ref{fig_cop_pedag1} are principally 
related in the left (respectively, right) portions of their distributions, a difference that may
have ecological significance, as discussed below, but is not captured by standard
correlation. Panels C and D 
were obtained from A and B through monotonic transformation (Appendix X for details),
<!--***DAN: Dan to go back and write the appendix-->
which alters the Pearson, but not the Spearman correlation. Although correlation
is only one of the numerous metrics for 
studying relationships between variables, other commonly used metrics also 
provide partial descriptions only.

Well-developed statistical tools do exist (REFS), and are applied widely in other fields 
(REFS), that provide a complete description of the dependence between two or more variables. 
The goal of this paper is to illustrate the potential for broader application of these
tools in ecology and related fields. The tools are based on the statistical concept of the *copula*, which we here introduce conceptually.

  a) then do it. some things to do include...
  b) the rank-based approach for data (use bivariate data)
  c) the inverse cdf approach for the copula of a bivariate distribution
  d) statement that copulas are also multivariate, though we won't explore that here (or maybe leave that for Discussion?)
  e) a one- or two-sentence description of sklars theorem

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-F, I, J) Diverse bivariate datasets, many of which have the same Pearson (P)
or Spearman (S) correlation coefficients. Panels C and E (respectively, D and F) were 
obtained from A (respectively, B) via monotonic functions that transformed the standard-
normal marginal distributions of A and B to gamma (C, D) or uniform (E, F) marginals. 
Details in Appendix X.\label{fig_cop_pedag1}}
\end{center}
\end{figure}
<!--***DAN: Dan to go back and add an appendix with details as it says. Make sure
to mention the shape and scale params of the gamma marginals in C, D were 2-->

@@

We used datasets that pertain to multiple fields and levels of organization within or related to ecology, including environmental, species-trait, phenological, population, community, and ecosystem functioning data. Copula structure describing the relationship between atmospheric weather variables such as X and Y has been examined previously in the meteorological literature (e.g., REF). We therefore examined environmental variables from the soil instead, using the Rapid Carbon Assessment database (RaCA; @Wills2014).
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil nitrogen (Mg C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. X for sampling locations).
<!--***DAN: Shya, please add a sup mat figure showing the sampling locations and a link to it here in place of the X
***Shya: once Prof. Loecke give me the lattitude and longitudes, I will do that.-->
Measurments were from the central pedon of a soil core, to 1 m depth.
<!--***DAN: Terry to check the wording and accuracy here-->
Species-trait data were average species basal metabolic rate (KJ per hour) and body mass (gram) data for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years from 1976 to 2010. These time series were computed from the the Rothamsted Insect 
Survey suction-trap dataset [@Macaulay1988, @Harrington2014]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years; these are 
indicators of flying aphid abundance. Data pre-processing for the aphid data was the same as 
that of @Sheppard2016,
but we exlcuded one of the 11 sites they used so as to include only sites with at least 30 
years data.  Our second population dataset comprised average annual abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 plankton taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years from 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. These data 
were obtained from the Continuous Plankton Recorder (CPR) dataset maintained by the Sir 
Alistair Hardy Foundation for Ocean Science (SAHFOS), now part of the Marine Biological 
Association of the UK. Data pre-processing was similar but not identical to that 
of @Sheppard2017; see Appendix \ref{SM-Data_details} for details.
Community-level data were plant aboveground biomass and Shannon's diversity index for X 
<!--***DAN: Shya, please fill in X with an autolink-->
plots, each 9m by 9m, maintained at Cedar Creek Ecosystem Science Reserve and sampled in 
the year 2000 [@Tilman2001; @Tilman2006]. 
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016]. Each included location was measured on at least 50 dates. 
Sampling locations spanned a hydrologic gradient from upland to seasonal wetland.
See Appendix \ref{SM-Data_details} for additional data details.

@@

# Concepts for Q1: Statistical tools for detecting copula structure \label{Methods}

Question Q1 from the Introduction was answered via statistical analysis of the
data described above, using both model selection and nonparametric approaches.
